A Survey of Quantitative Descriptions of Molecular Structure

### 2012-04-08 paper.tex
More etext of shape descriptors

While molecular surfaces are useful, molecular shape is a key factor
in ligand-receptor interactions. A variety of descriptors have been
devised to characterize molecular shape. The use of overlapping atom
centered Gaussians to represent molecular shape \cite{Grant:1995aa}
represents a significant computational advantage over the traditional
analytical approach of intersecting hard spheres, supporting extremely
rapid shape comparisons \cite{Grant:1999vn}. An alternative approach
described by Ritchie et al \cite{Ritchie:1999kx} employed a spherical
harmonic based representation of surfaces, that in contrast to the
Gaussian approach allows one to generate characterizations of the
surface and hence the shape at varying resolutions. In addition to
functional representations of molecular shape, a number of simpler
formulations have been defined. For example, the ``shape signature''
approach described by Zauhar et al \cite{Zauhar:2003fk} is a novel
application of a ray tracing method that generates a description of
the molecular shape. Ballester et al \cite{Ballester:2007aa} described
the Ultrafast Shape Recognition (USR) approach that characterized
molecular shape in terms of the distribution of distances between the
atoms and a set of pre-defined points. As a result, the calculation
procedure is extremely fast and once calculated shape similarities can
be evaluated rapidly via a simple distance calculation.

@@

>>>>

While molecular surfaces are useful, molecular shape is a key factor
in ligand-receptor interactions. A variety of descriptors have been
devised to characterize molecular shape. The use of overlapping atom
centered Gaussians to represent molecular shape \cite{Grant:1995aa}
represents a significant computational advantage over the traditional
analytical approach of intersecting hard spheres, supporting extremely
rapid shape comparisons \cite{Grant:1999vn}. An alternative approach
described by Ritchie et al \cite{Ritchie:1999kx} employed a spherical
harmonic based representation of surfaces, that in contrast to the
Gaussian approach allows one to generate characterizations of the
surface and hence the shape at varying resolutions. Another approach
to multi-resolution shape representations is the use of $\alpha$
shapes \cite{Edelsbrunner:1994aa}, which represent a parametrization
of the convex hull derived from the molecular coordinates. This
approach has been employed to characterize shape and volume
\cite{Wilson:2009ys} though it is more computationally intensive than
the methods mentioned previously. In addition to functional
representations of molecular shape, a number of simpler formulations
have been defined. For example, the ``shape signature'' approach
described by Zauhar et al \cite{Zauhar:2003fk} is a novel application
of a ray tracing method that generates a description of the molecular
shape. Ballester et al \cite{Ballester:2007aa} described the Ultrafast
Shape Recognition (USR) approach that characterized molecular shape in
terms of the distribution of distances between the atoms and a set of
pre-defined points. This procedure characterizes the shape in the form
of a 12-D real valued fingerprint. As a result, the calculation
procedure is extremely fast and once calculated shape similarities can
be evaluated rapidly via a simple distance calculation.

@@

Finally, quantum mechanical calculations can also generated
descriptors that can be applied to problems such as QSAR models
(though of course the time complexity of QM derived descriptors do not
make them amenable for large scale virtual screening applications).
